---
title: "Biomed case study 6"
author: "D. Anadria, H. Lee, L. Roos"
output: html_document
---
```{r}
# libraries
library(foreign)
library(tableone)
library(ggplot2)
library(tidyr)
library(ggpubr)
library(car)
library(stats)
library(pwr)
```

```{r}
# load data
choldat <- as.data.frame(read.spss(file = "data_cholesterol.sav"))
```
# Create table 1
```{r}
colnames_full <- as.list(colnames(choldat))
cont <- colnames_full[- c(1, 11, 12, 14, 15, 17, 18, 19, 20)]
fact <- colnames_full[c(5,6,7)]
                            
tab1 <-  CreateTableOne(vars = as.character(cont), 
                        factorVars = as.character(fact),
                        strata = "code",
                        data = choldat,
                        test = FALSE)
print(tab1, showAllLevels = TRUE)
```
```{r}
tab1Mat <- print(tab1, 
                 showAllLevels = TRUE, 
                 quote = FALSE, 
                 printToggle = FALSE)
# Save to a CSV file
write.csv(tab1Mat, file = "tab1.csv")
```

# Sample size 
```{r}
pwr.anova.test(k =2, power=0.8, f=2/15, sig.level =0.05)
```
```{r}
n <- 222*2
n * 1.05
```

# ANCOVA 
```{r}
ancov_res <- glm(HDLchol24 ~ code + HDLchol0, 
            data=choldat)
summary(ancov_res)
```
```{r}
ancov_res2 <- aov(HDLchol24 ~ code + HDLchol0, 
            data=choldat)
summary(ancov_res2)
```
```{r}
ancov_res2$coefficients
```
```{r}
confint(ancov_res2)
```

```{r}
choldat %>% 
  group_by(code) %>% 
  get_summary_stats(HDLchol24, type = "mean_sd")
```

# Plot
```{r}
choldat_long <- pivot_longer(data = choldat,
             cols = c("HDLchol0", "HDLchol12", "HDLchol24"),
             names_prefix = "HDLchol",
             names_to = "time",
             values_to = "HDLchol")

choldat_long %>%
  group_by(code, time) %>%
  get_summary_stats(HDLchol, type = "mean_sd")

choldat_long2 <- drop_na(choldat_long, "HDLchol")

boxplot <- ggboxplot(
  choldat_long2, 
  x = "code", 
  y = "HDLchol", 
  color = "time",
  xlab = 'Group',
  main = "Figure 1: Boxplots per time point per group"
  )
boxplot
```
# Assumption checking

## Linearity
```{r}
scatterplot(choldat$HDLchol24 ~ choldat$HDLchol0 | code, 
            data = choldat, 
            grid = FALSE,
            col = c("red","blue"),
            pch = c(5,1),
            main = "Linearity between covariate and outcome (per group)",
            xlab = "Baseline HDL",
            ylab = "Final HDL")
```
## Homogeneity of regression slopes
```{r}
summary(aov(HDLchol24 ~ code*HDLchol0, 
    data = choldat))
```
Test is significant but looking at above plot, we can assume homogeneity. 

## Outliers
```{r}
boxplot(choldat$HDLchol24)
```

## Normality of residuals

```{r}
shapiro.test(choldat$HDLchol24)
```

Q-Q plot
```{r}
qqnorm(residuals(ancov_res))
qqline(residuals(ancov_res)) 
```

```{r}
# X-axis grid
x <- choldat$HDLchol24
x2 <- seq(min(x, na.rm=T), max(x, na.rm=T), length = 40)

# Normal curve
fun <- dnorm(x2, mean = mean(x, na.rm=T), sd = sd(x, na.rm=T))

# Histogram
hist(x, prob = TRUE,
     main = "Density of the data versus normal curve")
lines(x2, fun, col = 2, lwd = 2)
lines(density(x, na.rm=T), col = 4, lwd = 2)
```


## Homogeneity of variance / Homoscedasticity  
```{r}
leveneTest(HDLchol24 ~ as.factor(code), data = choldat)
```
```{r}
var.test(HDLchol24 ~ as.factor(code), data = choldat)
```
```{r}
fligner.test(HDLchol24 ~ as.factor(code), data = choldat)
```
```{r}
plot(fitted(ancov_res), residuals(ancov_res))
abline(0,0)
```
So, heteroscedasticity...

# Outliers removed
```{r}
outliers24 <- boxplot.stats(choldat$HDLchol24)$out
out_rows24 <- which(choldat$HDLchol24 %in% c(outliers24))

outliers0 <- boxplot.stats(choldat$HDLchol0)$out
out_rows0 <- which(choldat$HDLchol0 %in% c(outliers0))

outliers <- unique(c(out_rows24, out_rows0))
# remove outliers
choldat_new <- choldat[-c(outliers),]
```

## ANCOVA 
```{r}
ancov_res_out <- glm(HDLchol24 ~ code + HDLchol0, 
            data=choldat_new)
summary(ancov_res_out)
```

```{r}
ancov_res_out2 <- aov(HDLchol24 ~ code + HDLchol0, 
            data=choldat_new)
summary(ancov_res_out2)
```
```{r}
ancov_res_out2$coefficients
confint(ancov_res_out2)
```
```{r}
choldat_new %>% 
  group_by(code) %>% 
  get_summary_stats(HDLchol24)
```
```{r}
# r-squared
with(summary(ancov_res_out), 1 - deviance/null.deviance)
```

## Plot
```{r}
choldat_long <- pivot_longer(data = choldat_new,
             cols = c("HDLchol0", "HDLchol12", "HDLchol24"),
             names_prefix = "HDLchol",
             names_to = "time",
             values_to = "HDLchol")

choldat_long %>%
  group_by(code, time) %>%
  get_summary_stats(HDLchol, type = "mean_sd")

choldat_long2 <- drop_na(choldat_long, "HDLchol")

boxplot <- ggboxplot(
  choldat_long2, 
  x = "code", 
  y = "HDLchol", 
  color = "time",
  xlab = 'Group',
  main ="Figure 1: HDL levels per time point and per group"
  )
boxplot
```

## Assumption checking

### Linearity
```{r}
scatterplot(choldat_new$HDLchol24 ~ choldat_new$HDLchol0 | code, 
            data = choldat_new, 
            grid = FALSE,
            col = c("red","blue"),
            pch = c(5,1),
            main = "Linearity between covariate and outcome (per group)",
            xlab = "Baseline HDL",
            ylab = "Final HDL")
```

## Homogeneity of regression slopes
```{r}
summary(aov(HDLchol24 ~ code*HDLchol0, 
    data = choldat_new))
```

## Normality of residuals

```{r}
shapiro.test(choldat_new$HDLchol24)
```

Q-Q plot
```{r}
qqnorm(residuals(ancov_res_out))
qqline(residuals(ancov_res_out)) 
```

```{r}
# X-axis grid
x <- choldat_new$HDLchol24
x2 <- seq(min(x, na.rm=T), max(x, na.rm=T), length = 40)

# Normal curve
fun <- dnorm(x2, mean = mean(x, na.rm=T), sd = sd(x, na.rm=T))

# Histogram
hist(x, prob = TRUE,
     main = "Density of the data versus normal curve")
lines(x2, fun, col = 2, lwd = 2)
lines(density(x, na.rm=T), col = 4, lwd = 2)
```

## Homogeneity of variance / Homoscedasticity  
```{r}
leveneTest(HDLchol24 ~ as.factor(code), data = choldat_new)
```
```{r}
var.test(HDLchol24 ~ as.factor(code), data = choldat_new)
```
```{r}
fligner.test(HDLchol24 ~ as.factor(code), data = choldat_new)
```
```{r}
plot(fitted(ancov_res_out), residuals(ancov_res_out))
abline(0,0)
```
Still a little bit of heterscedasticity, but nothing to worry about. 